RED WINE QUALITY EXPLORATION by FELIPE BARROS
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(gridExtra)
library(dplyr)
library(RColorBrewer)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
setwd("C:/Users/flfbarros/Documents/udacity//R/Red Wine Exploration")
redwine <- read.csv("wineQualityReds.csv", sep = ",")
```

This dataset is one of the datasets created on the paper "Modeling wine 
preferences by data mining from physicochemical properties" [Cortez et al., 
2009]. The inputs include objective tests, such as PH values, and the output is 
based on sensory data, gathered from at least 3 wine expert evaluations. The 
output value ranges from 0 (very bad) to 10 (very excellent). The wines are 
variants of the Portuguese "Vinho Verde" wine. This particular dataset contains 
information regarding the red wines.

There are 1599 different wines, 11 features and 1 output value (quality).
Information about the features and the output value is the following (provided 
by the original authors):

Input variables (based on physicochemical tests):

* 1 - fixed acidity (tartaric acid - g / dm^3)
* 2 - volatile acidity (acetic acid - g / dm^3)
* 3 - citric acid (g / dm^3)
* 4 - residual sugar (g / dm^3)
* 5 - chlorides (sodium chloride - g / dm^3
* 6 - free sulfur dioxide (mg / dm^3)
* 7 - total sulfur dioxide (mg / dm^3)
* 8 - density (g / cm^3)
* 9 - pH
* 10 - sulphates (potassium sulphate - g / dm3)
* 11 - alcohol (% by volume)

Output variable (based on sensory data):

* 12 - quality (score between 0 and 10)

Description of the features and output value is the following (also provided 
by the original authors):

* 1 - fixed acidity: most acids involved with wine or fixed or nonvolatile 
(do not evaporate readily)
* 2 - volatile acidity: the amount of acetic acid in wine, which at too high 
of levels can lead to an unpleasant, vinegar taste
* 3 - citric acid: found in small quantities, citric acid can add 'freshness' 
and flavor to wines
* 4 - residual sugar: the amount of sugar remaining after fermentation stops, 
it's rare to find wines with less than 1 gram/liter and wines with greater 
than 45 grams/liter are considered sweet
* 5 - chlorides: the amount of salt in the wine
* 6 - free sulfur dioxide: the free form of SO2 exists in equilibrium between 
molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial 
growth and the oxidation of wine
* 7 - total sulfur dioxide: amount of free and bound forms of S02; in low 
concentrations, SO2 is mostly undetectable in wine, but at free SO2 
concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine
* 8 - density: the density of water is close to that of water depending on the 
percent alcohol and sugar content
* 9 - pH: describes how acidic or basic a wine is on a scale from 0 (very 
acidic) to 14 (very basic); most wines are between 3-4 on the pH scale
* 10 - sulphates: a wine additive which can contribute to sulfur dioxide gas 
(S02) levels, wich acts as an antimicrobial and antioxidant
* 11 - alcohol: the percent alcohol content of the wine
* 12 - quality (score between 0 and 10)

# Univariate Plots Section

```{r echo=FALSE, message=FALSE}

str(redwine)

```


```{r echo=FALSE, Univariate_Plots}

summary(subset(redwine, select = -c(X)))

```

The dataset consists of 11 numerical features and 1 integer feature (quality). 
The X feature just represents the index of the observation on the data frame.

On the summary it is possible to identify some initial trends on the data. For 
example, there are some features that seem to have outliers on them, such as: 
volatile acidity, citric acid, redisual sugar, chlorides, free sulfur dioxide, 
total sulfur dioxide, and sulphates. Also, most of the wines are of average 
quality (the 1st quartile border equals 5, the median equals 6 and the third 
quartile border equals 6).

```{r echo=FALSE, message=FALSE, warning=FALSE}

table(redwine$quality)

```

Out of 1599 wines, 63 (~4%) are of quality 3 or 4, 1319 (~82.5%) are of quality 
5 or 6 and 217 (~13,6%) are of quality 7 or 8.

Histograms and boxplots will help visualize distribution patterns for the other 
features as well.

```{r echo=FALSE, message=FALSE, warning=FALSE}
hg <- ggplot(data = redwine, aes(x = quality)) +
  geom_histogram()

bp <- ggplot(data = redwine, aes(y = quality, x = 1)) +
  geom_jitter(alpha = 0.25) +
  geom_boxplot(alpha = 0.5, color = "blue")

grid.arrange(hg, bp, ncol=2)


```

We can easily see the difference between number of observations for the values 
of quality. Despite that, the distribution is normal and there aren't a lot of 
outliers on the feature.

```{r echo=FALSE, message=FALSE, warning=FALSE}
hg <- ggplot(data = redwine, aes(x = fixed.acidity)) +
  geom_histogram()

bp <- ggplot(data = redwine, aes(y = fixed.acidity, x = 1)) +
  geom_jitter(alpha = 0.25) +
  geom_boxplot(alpha = 0.5, color = "blue")

grid.arrange(hg, bp, ncol = 2)

```

Fixed acidity has a more sparse distribution, although still normal-like. There 
are a higher number of outliers on the higher end, however. These could either 
be removed or omitted on future plots.

```{r echo=FALSE, message=FALSE, warning=FALSE}

hg <- ggplot(data = redwine, aes(x = volatile.acidity)) +
  geom_histogram()

bp <- ggplot(data = redwine, aes(y = volatile.acidity, x = 1)) +
  geom_jitter(alpha = 0.25) +
  geom_boxplot(alpha = 0.5, color = "blue")

grid.arrange(hg, bp, ncol = 2)

```

Volatile acidity has a similar distribution as fixed acidity, but with less 
outliers on the higher end. Again, this could be either removed completely or 
omitted on future plots.

```{r echo=FALSE, message=FALSE, warning=FALSE}

hg <- ggplot(data = redwine, aes(x = citric.acid)) +
  geom_histogram()

bp <- ggplot(data = redwine, aes(y = citric.acid, x = 1)) +
  geom_jitter(alpha = 0.25) +
  geom_boxplot(alpha = 0.5, color = "blue")

grid.arrange(hg, bp, ncol = 2)

```

Citric acid has the most disperse distribution so far. The boxplot sections are 
larger and there are almost no outliers. However, the data does not follow a 
normal distribution.

```{r echo=FALSE, message=FALSE, warning=FALSE}

hg <- ggplot(data = redwine, aes(x = residual.sugar)) +
  geom_histogram()

bp <- ggplot(data = redwine, aes(y = residual.sugar, x = 1)) +
  geom_jitter(alpha = 0.25) +
  geom_boxplot(alpha = 0.5, color = "blue")

grid.arrange(hg, bp, ncol = 2)

```

Residual sugar has a very large number of outliers. So much that it is hard to 
do a proper analysis on these plots. To solve this, we could either transform 
the feature with a log of square root transformation, and/or remove or omit the 
outliers.

```{r echo=FALSE, message=FALSE, warning=FALSE}

hg <- ggplot(data = redwine, aes(x = residual.sugar)) +
  geom_histogram() +
  scale_x_log10()

bp <- ggplot(data = redwine, aes(y = residual.sugar, x = 1)) +
  geom_jitter(alpha = 0.25) +
  geom_boxplot(alpha = 0.5, color = "blue") +
  scale_y_log10()

grid.arrange(hg, bp, ncol = 2)

```

With a log transformation, it is easier to understand the distribution of the 
feature, but there's still a lot of outliers difficulting the process.

```{r echo=FALSE, message=FALSE, warning=FALSE}

hg <- ggplot(data = redwine, aes(x = residual.sugar)) +
  geom_histogram() +
  scale_x_log10() +
  scale_x_continuous(limits = c(0.5,
                                quantile(redwine$residual.sugar, 0.9)))

bp <- ggplot(data = redwine, aes(y = residual.sugar, x = 1)) +
  geom_jitter(alpha = 0.25) +
  geom_boxplot(alpha = 0.5, color = "blue") +
  scale_y_log10() +
  scale_y_continuous(limits = c(0.5,
                                quantile(redwine$residual.sugar, 0.9)))

grid.arrange(hg, bp, ncol = 2)

```

By doing a log transformation and removing the top 10%, the information is even 
clearer. The distribution doesn't follow a normal distribution (there are 
peaks on two different sections). The feature so far is the most affected by 
outliers.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = redwine, aes(x = chlorides)) +
  geom_histogram()

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = redwine, aes(x = free.sulfur.dioxide)) +
  geom_histogram()

```

```{r  echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = redwine, aes(x = total.sulfur.dioxide)) +
  geom_histogram()

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = redwine, aes(x = density)) +
  geom_histogram()

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = redwine, aes(x = pH)) +
  geom_histogram()

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = redwine, aes(x = sulphates)) +
  geom_histogram()

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = redwine, aes(x = alcohol)) +
  geom_histogram()

```

These histograms help us visualize how some of the features on the dataset have 
outliers, given that some of them are right tailed. Chlorides is a good example 
of this characteristic. Most of the data is around 0.05 and 0.15, with very 
few examples of wine going up to 6.

Also, most of the data is relatively normal, with the exception of citric acid.

Let's transform some of the features using a log scale to try and get better 
visualizations of them. Suitable choices would be residual sugar, chlorides, 
free sulfur dioxide, total sulfur dioxide, and sulphates.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = redwine, aes(x = residual.sugar)) +
  geom_histogram() +
  scale_x_log10()

```


```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x = chlorides)) +
  geom_histogram() +
  scale_x_log10()

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = redwine, aes(x = free.sulfur.dioxide)) +
  geom_histogram() +
  scale_x_log10()

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = redwine, aes(x = total.sulfur.dioxide)) +
  geom_histogram() +
  scale_x_log10()

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = redwine, aes(x = sulphates)) +
  geom_histogram() +
  scale_x_log10()

```


With log transformation it is easier to identify patterns on these features. 
With the exception of free sulfur dioxide, they all have normal-like 
distributions.

For some of the features it might be useful to divide them into buckets. For 
quality, reducing the various possible values into buckets helps plot 
visualization and understanding how the values fare against each other.
Features like alcohol and density can naturally describe properties of the 
wines in a easy to understand way (e.g. lighter, more alcoholic, etc.). The 
three features will be divided into buckets and be used in both ways 
(continuous and categorical) onwards.

For quality, the baskets are:

* Bad: 4 or lower quality
* Average: 5 quality
* Good: 6 quality
* Best: 7 or over quality

The choice of breaks for the baskets are based on the fact that most 
observations on the dataset have quality of 5 or 6.

```{r echo=FALSE, warning=FALSE, message=FALSE}

redwine$quality_factor <- cut(redwine$quality, breaks = c(0,4,5,6,10), 
                              labels = c("Bad", "Average", "Good", "Best"))

```

```{r echo=FALSE, warning=FALSE, message=FALSE}

ggplot(data = redwine, aes(x = quality_factor)) +
  geom_bar()

```

Since most of the wines have quality of 5 or 6, most of them are being 
described here as "Average" or "Good", with a good ammount of "Best".

The value range for alcohol is of 6.5 (14.9 minus 8.4). By dividing alcohol 
levels into three buckets, that would imply in a step of around 2.16 for each 
bucket. I'll round it to 2.15 and use that value as a step to the cut. The 
buckets will be cut as follows:

* Low alcohol level: over 8.4% or under or equal to 9.55%
* Medium alcohol level: over 9.55% or under or equal to 11.7%
* High alcohol level: over 11.7%

```{r echo=FALSE, message=FALSE, warning=FALSE}

redwine$alcohol_level <- cut(redwine$alcohol, breaks = c(8,9.55,11.7,15), 
                             labels = c("Low Alcohol",
                                        "Medium Alcohol",
                                        "High Alcohol"))

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = redwine, aes(x = alcohol_level)) +
  geom_bar()

```

Most wines are bucketed into Medium Alcohol, followed by Low Alcohol and 
High Alcohol. Although there is a big difference in the number of observations 
between the buckets, they all have a reasonable number to investigate.

The variance in density levels is 0.0136. By diving in three equal steps, we've 
got roughly 0.0045 for each step. However, the first step doesn't even surpass 
the first quartile, and the second goes from there to beyond the third. That 
would create a bucket with the majority of the observations. Looking to create 
more even buckets, I'll use the values from the first quartile, median and 
third quartile.

```{r echo=FALSE, message=FALSE, warning=FALSE}

redwine$density_level <- cut(redwine$density, 
                             breaks = c(0, 0.9956, 0.9978, 1.0038), 
                             labels = c("Light", "Medium", "Heavy"))

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = redwine, aes(x = density_level)) +
  geom_bar()

```

The distribution seems more balanced, even with the middle bucket with almost 
double the number of observations of the other buckets. The buckets are 
represented in the following manner:

* Light wines: up to 0.9956g/cm^3
* Medium wines: higher than 0.9956g/cm^3 and up to 0.9978g/cm^3
* Heavy wines: higher than 0.9978g/cm^3

# Univariate Analysis

The dataset consists of 1599 observations of red wines with information about 
11 physicochemical properties and 1 information about the quality of each wine. 
The pyhisicochemial information was collected with experiments and the quality 
of each wine is a mean of a value provided by different wine experts. All the 
information is in the form of numbers.

The quality values were given in a scale from 0 (very bad) to 10 (very 
excellent). However, the majority of the wines have quality of 5 or 6. The 
worst wine has a quality of 3. The best wine has a quality of 8.

Some features are long-tailed. Examples are residual sugar, chlorides, free 
sulfur dioxide, total sulfur dioxide, and sulphates.

With a few exceptions, the features have a distribution close to normal.

Since quality is the output feature of the study that generated this dataset, 
the feature is undoubtly the main feature of interest.

For the variance observed in some of these features and their nature in 
relation to overall taste in wine, I would describe residual sugar, density and 
pH as features of interest.

So far, I did not create any new variables from existing variables in the 
dataset.

# Bivariate Plots Section

I've established some features of interest, but that was mainly done looking at 
their distribution and common wine knowledge. It would be best to evaluate the 
correlation between the features, to identify if the choices made are the best 
ones or if there is some possible feature worth considering.

```{r echo=FALSE, Bivariate_Plots}

cor(subset(redwine, select = -c(X)))

```

Based on correlations, the best features to predict the quality of a wine are, 
in order, alcohol, volatile acidity and sulphates. My initial intuition was 
that residual sugar, density and pH would better describe the quality of a 
wine, with alcohol and acidity features as possible features to look afterwards. 
These initial findings show that there is more to the quality of a wine than 
what common knowledge says.

With the correlation knowledge, I'll try now to visualize the influence of 
those features on the quality of the wines.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = redwine, aes(y = alcohol, x = quality)) +
  geom_point(position = position_jitter(width = 0.2, height = 0.2)) +
  geom_smooth(method='lm')

```

The scatterplot suggests that as the percentage of alcohol increases so does 
the overall quality of the wine, as shown by the linear regression line. Even 
when comparing quality of 5 and 6 (the most common on the dataset), the 
observations tend to go up on quality 6, whilst being clumped below the 
regression line on quality 5.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = redwine, aes(y = alcohol, x = as.factor(quality))) +
  geom_jitter(alpha = 0.25) +
  geom_boxplot(alpha = 0.5, color = "blue") +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               shape = 3, 
               color = "red", 
               size = 4) +
  xlab("quality")

```

The boxplot also supports that theory. Wines with better quality have higher 
median and higher mean (shown as red crosses).

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = redwine, aes(y = volatile.acidity, x = quality)) +
  geom_point(position = position_jitter(width = 0.2, height = 0.1)) +
  geom_smooth(method='lm')

```

Volatile acidity is the opposite as alcohol: as it increases, the overall 
quality of the wine decreases. Since the correlation found was negative 
(around -0.39), that was expected.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = redwine, aes(x = as.factor(quality), y = volatile.acidity)) +
  geom_jitter(alpha = 0.25) +
  geom_boxplot(alpha = 0.5, color = "blue") +
  stat_summary(fun.y = "mean",
               geom = "point",
               shape = 3,
               color = "red",
               size = 4) +
  xlab("quality")

```

Again, the boxplots suggests the same and provide additional information (mean
and median values for each quality). As the level of volatile acidity decreases,
the perceived quality of the wine increases.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = redwine, aes(y = sulphates, x = quality)) +
  geom_point(position = position_jitter(width = 0.2, height = 0.05)) +
  geom_smooth(method = 'lm') +
  scale_y_continuous(limits = c(0.25, quantile(redwine$sulphates, 0.99)))

```

The top 1% for sulphates was removed for better visualization. For sulphates, 
the scatterplot has a behavior similar to alcohol: better quality wines have 
higher levels of sulphates.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = redwine, aes(x = as.factor(quality), y = sulphates)) +
  geom_jitter(alpha = 0.25) +
  geom_boxplot(alpha = 0.5, color = "blue") +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               shape = 3, 
               color = "red",
               size = 4) +
  xlab("quality")

```

The boxplot for each quality supports that trend. All the higher qualities have 
a higher mean and median values of sulphates.

By plotting these three features on scatterplots and boxplots against quality, 
it is easier to visualize their influence on the quality of the wines. By now, 
it is safe to say that better wines probably have higher alcohol and sulphates 
levels, whilst having a lower volatile acidity level.

Since citric acid is the fourth most correlated feature to quality, it is worth 
doing some analysis on it.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = redwine, aes(y = citric.acid, x = quality)) +
  geom_point(position = position_jitter(width = 0.2, height = 0.1)) +
  geom_smooth(method = 'lm')

```

The correlation between citric acid and quality (~0.226) is not much different 
from the correlation between sulphates and quality (~0.25), so their tendencies 
are also similar to each other. Better wines tend to have, on average, higher 
levels of citric acid.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = redwine, aes(x = as.factor(quality), y = citric.acid)) +
  geom_jitter(alpha = 0.25) +
  geom_boxplot(alpha = 0.5, color = "blue") +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               shape = 3, 
               color = "red",
               size = 4) +
  xlab("quality")

```

For the boxplot between citric acid and quality, we can see a clear difference
between better wines (quality of 7 and 8) and the other ones. Not only they 
tend to have higher values of citric acid, the majority of their observations 
(over 75%) are over 0.25g/dm^3 of citric acid.

Since density is a feature most commonly used when describing "the body" of the 
wine, I'll do a similar analysis for it.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = redwine, aes(x = quality, y = density)) +
  geom_point(position = position_jitter(width = 0.2, height = 0.01)) +
  geom_smooth(method = 'lm') +
  scale_y_continuous(limits = c(0.99, quantile(redwine$density, 0.99)))

```

The top 1% for density was removed for better visualization. Since the 
correlation between these features is very low (around -0.17), the influence of 
density on the quality of the wine is lower than previous features. It seems 
that using density to evaluate the quality of a wine (a common practice of 
non-experts) is not the best approach.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = redwine, aes(x = as.factor(quality), y = density)) +
  geom_jitter(alpha = 0.25) +
  geom_boxplot(alpha = 0.5, color = "blue") +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               shape = 3, 
               color = "red",
               size = 4) +
  xlab("quality")

```

The boxplot adds a little more information. There is a decrease of density 
levels on better wines, but that is only noticeable with quality levels 
of 7 and 8. For the other quality levels, there doesn't seem to be a 
significant difference.

# Bivariate Analysis

I've investigated the relationships between quality and alcohol, volatile 
acidity, sulphates, citric acid and density. In terms of correlation with 
quality, they were, in order, 1st, 2nd, 3rd, 4th and 6th most correlated 
features of the dataset. The first four were chosen because of they closer 
relationship with quality and density was chosen out of personal curiosity.

Alcohol was the feature most correlated with quality, with a factor of 0.48. 
The positive correlation between them implied that increasing the level of 
alcohol would increase the perceived quality of the wine. The plots created 
demonstrated that assumption.

Volatile acidity was the second feature most correlated with quality, with a 
factor of -0.39. The negative correlation implies a decrease of one when an 
increase in the other occurs. The plots generated help visualize that 
relationship. As volatile acidity levels go down, the quality of the wines go 
up.

Sulphates and citric acid where the third and fourth most correlated qualities, 
with factors of 0.25 and 0.226, respectively. Higher levels of these features 
are correlated with better quality of the wine, although their influence is not 
as strong as the influence of alcohol and volatile acidity.

Density was the sixth most correlated feature with quality. The choice to 
visualize the influence of that feature was mostly curiosity based, since the 
correlation factor is of only -0.175. Although there is an influence of the 
feature on the quality of the wine, that influence is only noteworthy on the 
best wines on the dataset (7 or above). Those wines tend to have a lower 
density value.

Features related to acidity naturally seem to be correlated with each other. 
Examples are fixed acidity, volatile acidity, citric acid and pH. Density also 
seems to be related with those features, along with residual sugar and alcohol.

The strongest relationship found was between fixed acidity and pH, with a 
factor of 0.683. That is not a surprising relationship, as one feature (pH) is 
evaluated based on properties of the other (acidity).

# Multivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE}
#I decided to create this custom palette to use on my plots
#source: https://stackoverflow.com/questions/29466239/how-to-set-the-color-range-of-scale-colour-brewer-in-ggplot2-palette-selecte
my_palette <- brewer.pal(n = 9, "Greens")[4:7]

```


```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = redwine, aes(x = alcohol, y = volatile.acidity, 
                         color = quality_factor)) +
  geom_point(position = position_jitter(width = 0.2, height = 0.01),
             alpha = 0.75,
             size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, size = 1.1) +
  scale_y_continuous(breaks = seq(0, 1.5, 0.1), 
                     limits = c(0, 
                                quantile(redwine$volatile.acidity, 0.99))) +
  geom_vline(xintercept = median(redwine$alcohol), 
             linetype = "longdash") +
  geom_hline(yintercept = median(redwine$volatile.acidity), 
             linetype = "dashed") +
  scale_color_manual(values = my_palette)

```

The top 1% values for volatile acidity were removed on this plot. The vertical 
and horizontal lines are the medians for alcohol and volatile acidity, 
respectively. By using both lines, the graph is divided into four quadrants. 
We can see how alcohol has a higher influence in the quality factor as most of 
bad and average wines are on the second or third quadrants, where alcohol 
levels are lower. The first quadrant has a good number of good wines but also 
has some presence of bad and average wines. The fourth quadrant, where alcohol 
is high and volatile acidity is low is the best quadrant, as expected: most 
best wines are present there.


```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = redwine, aes(x = alcohol, y = density, color = quality_factor)) +
  geom_point(position = position_jitter(width = 0.2, height = 0.0025),
             alpha = 0.75,
             size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, size = 1.1) +
  geom_hline(yintercept = median(redwine$density), 
             linetype = "longdash") +
  geom_vline(xintercept = median(redwine$alcohol), 
             linetype = "longdash") +
  scale_color_manual(values = my_palette)

```

By swaping volatile acidity for density on the y-axis, the pattern is similar, 
but with even more bias towards the x-axis (alcohol). The vertical line (the 
median for alcohol) seems to be the only one dividing most of the good ones 
from the bad ones. This again goes against common knowledge that density is 
a good feature for telling the quality of a wine.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = redwine,
       aes(x = volatile.acidity, y = sulphates, color = quality_factor)) +
  geom_point(position = position_jitter(width = 0.01, height = 0.01),
             alpha = 0.75,
             size = 1) +
  geom_smooth(method = "lm", se = FALSE, size = 1.1) +
  scale_y_continuous(limits = c(0.25, 
                                quantile(redwine$sulphates, 0.99))) +
  scale_x_continuous(limits = c(0.1, 
                                quantile(redwine$volatile.acidity, 0.99))) +
  geom_hline(yintercept = median(redwine$sulphates), 
             linetype = "longdash") +
  geom_vline(xintercept = median(redwine$volatile.acidity), 
             linetype = "longdash") +
  scale_color_manual(values = my_palette)

```

This plot also has a difference between the quadrants. In this case, the second 
quadrant (high sulphates and low volatile acidity) is where most of the best 
wines are.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = redwine, 
       aes(x = sulphates, y = volatile.acidity, color = quality_factor)) +
  geom_point(alpha = 0.75,
             size = 1) +
  geom_smooth(method = "lm", se = FALSE, size = 1) +
  facet_wrap(~ alcohol_level) +
  scale_x_continuous(limits = c(0.25, 
                                quantile(redwine$sulphates, 0.99))) +
  scale_y_continuous(limits = c(0.125, 
                                quantile(redwine$volatile.acidity, 0.99))) +
  geom_hline(yintercept = median(redwine$volatile.acidity), 
             linetype = "longdash") +
  geom_vline(xintercept = median(redwine$sulphates), 
             linetype = "longdash") +
  scale_color_manual(values = my_palette)

```

The top 1% for both axes were removed for this plot, in order to better 
visualize it. By dividing the alcohol levels into buckets, we can see not only 
how the quality of the wines are distributed in the buckets but also the
distrubution of sulphates and volatile acidity. The high alcohol bucket has 
almost only good and best wines and most of the observations are on the fourth 
quadrant. On the opposite side, the low alcohol bucket has almost only average 
wines and most of the observations are on the second and third quadrants. The 
plot depicts at the same time the influence of the three features most 
impactful on the quality of a wine.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data = redwine, 
       aes(x = alcohol, y = volatile.acidity, color = quality_factor)) +
  geom_point(alpha = 0.75, size = 1) +
  geom_smooth(method = "lm", se = FALSE, size = 1) +
  facet_wrap(~ density_level) +
  scale_y_continuous(limits = c(0.25, 
                                quantile(redwine$volatile.acidity, 0.99))) +
  geom_vline(xintercept = median(redwine$alcohol), 
             linetype = "longdash") +
  geom_hline(yintercept = median(redwine$volatile.acidity), 
             linetype = "longdash") +
  scale_color_manual(values = my_palette)

```

Lower density levels are correlated with better wine. In the low density 
bucket, most of the wines are good or best, as expected. They also are higher
on alcohol levels. The medium density bucket is a surprise, as it has a more 
dense cloud of average wines than the high density bucket.

# Multivariate Analysis

By looking simultaneously at several features and their relationship with 
quality, it was possible to see how each other contribute to the quality of the 
wine. Since the three best features to try to predict the quality of the wine 
are alcohol, volatile acidity and sulphates, creating a plot with the 
relationship between them and quality efficiently described the properties of 
a good wine.

To improve the analysis, I decided to divide quality, alcohol and density 
into buckets. That allowed me to visualize multiple features interacting with 
each other at the same time.

A surprising interaction was found when density values were cut into buckets. 
Even though a linear regression model suggests that low density levels are 
correlated with better wine, the middle bucket seemed worst than the heavy 
bucket.

------

# Final Plots and Summary

```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}

ggplot(data = redwine, 
       aes(x = sulphates, y = volatile.acidity, color = quality_factor)) +
  geom_point(alpha = 0.75, size = 1) +
  geom_smooth(method = "lm", se = FALSE, size = 1) +
  facet_wrap(~ alcohol_level) +
  scale_x_continuous(limits = c(0.25, 
                                quantile(redwine$sulphates, 0.99))) +
  scale_y_continuous(limits = c(0.125, 
                                quantile(redwine$volatile.acidity, 0.99))) +
  geom_hline(yintercept = median(redwine$volatile.acidity), 
             linetype = "longdash") +
  geom_vline(xintercept = median(redwine$sulphates), 
             linetype = "longdash") +
  labs(
  title = "Wine Quality, based on Acetic Acid, Potassium Sulphate and Alcohol", 
  color = "Quality", 
  subtitle = "Low Alcohol: up to 9.55% 
Medium Alcohol: over 9.55% and up to 11.7% 
High Alcohol: over 11.7%"
  ) +
  ylab("Acetic Acid (g/dm^3)") +
  xlab("Potassium Sulphate (g/dm^3)") +
  scale_color_manual(values = my_palette)

```

This is the best plot to represent the dataset. We can see how alcohol heavily 
influences the quality of the wine by the difference of observations on each 
bucket. The median lines for Acetic Acid and Potassium Sulphate also help to 
illustrate how these features work in relation to the quality of the wine. More 
acetic acid gives a vinegar taste to the wine, so most wines above the 
horizontal line are Bad or Average. Potassium Sulphate helps with antimicrobial 
and antioxidant effects on wine, giving them an overall better quality. Because 
of that, wines on the right of the vertical line are more likely to be Good or 
Best.

```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}

ggplot(data = redwine, 
       aes(x = alcohol, y = volatile.acidity, color = quality_factor)) +
  geom_point(alpha = 0.75, size = 1) +
  geom_smooth(method = "lm", se = FALSE, size = 1) +
  facet_wrap(~ density_level) +
  scale_y_continuous(limits = c(0.25, 
                              quantile(redwine$volatile.acidity, 0.99))) +
  geom_vline(xintercept = median(redwine$alcohol), 
             linetype = "longdash") +
  geom_hline(yintercept = median(redwine$volatile.acidity), 
             linetype = "longdash") +
  labs(
    title = "Quality of a Wine based on Acetic Acid, Alcohol and Density", 
    subtitle = "Light: up to 0.9956g/dm^3
Medium: over 0.9956 and up to 0.9978g/dm^3
Heavy: over 0.9978g/dm^3", 
    color = "Quality"
    ) +
  xlab("Alcohol (%)") +
  ylab("Acetic Acid (g/dm^3)") +
  scale_color_manual(values = my_palette)

```

The second plot I've chosen to describe the data is the one above. The reason 
I've chosen it is because most people (myself included) are not experts on 
wine, and most of them believe that the density of the wine is somehow related 
to its quality. Although that is true, density has a positive correlation with 
the quality of the wine, the influence of density on the quality is low. 
Lighter wines are more likely to be Good or Best, but Medium and Heavy wines 
are more or less the same on that regard.

```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}

ggplot(data = redwine, aes(y = alcohol, x = as.factor(quality))) +
  geom_jitter(alpha = 0.25) +
  geom_boxplot(alpha = 0.5, color = "blue") +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               shape = 3, 
               color = "red",
               size = 4) +
  xlab("Quality") +
  labs(title = "Variance of Alcohol Levels by Quality of the Wine", 
       subtitle = "Crosses represent mean values") +
  ylab("Alcohol (%)")

```

These boxplots are a great description of the most influential feature on 
regarding to the quality of a wine: alcohol. With the strongest correlation 
with quality, variance in the levels of alcohol between wines is what best 
separates the worst wines from the best ones.

------

# Reflection

This was my first solo attempt at an Exploratory Data Analysis. The data used 
consisted of physicochemical properties and quality of wines. Since my 
knowledge of wine was close to none, I as mostly looking at the data without 
much assumptions. The only thing that I thought I knew was that people use 
the density of the wine to describe its "body", and that might be related to 
its quality.

As I went along with the exploration, I've discovered what features are more 
correlated with quality and if that correlation was positive or negative. I was 
surprised to discover that density was correlated, but not much. I had only one 
assumption and it turned out to be false!

The multivariate analysis was not easy to perform. Some of my first plots were 
not describing the data correctly, I used features on wrong axes, was not 
taking much out of them. Then, I decided to include the vertical and horizontal 
lines for the median values for each axis. After that, the information was 
clearer and the analysis was easier to make.

I've encountered two problems while doing this exploration. The first was that 
quality was the only naturally categorical feature. That limited my analysis 
at first, but I tried to go around that by creating two categorical features, 
derived from original features on the dataset. I think that, because of that, 
I was able to go further on my analysis. The second problem was the natural 
distribution of the data: most wines were of quality 5 or 6, with none below 3 
or above 8. That was not a problem I could solve in any way I can think of. 
It wasn't even a big problem, but it is necessary to address the results from 
this analysis with caution, as it could not be a good representation for other 
datasets of red wines.

Going further, some chemistry background knowledge could prove useful on trying 
to create more derived features. Maybe there is a relationship between the 
features on the dataset that I am not aware of. In reality, I've reserved most 
of the investigation to the three most correlated features and density (to 
address my only assumption). There could be much more hidden in the data I'm 
simply not yet experienced enough to unveil.
