RED WINE QUALITY EXPLORATION by FELIPE BARROS
========================================================

> **Tip**: You will see quoted sections like this throughout the template to
help you construct your report. Make sure that you remove these notes before
you finish and submit your project!

> **Tip**: One of the requirements of this project is that your code follows
good formatting techniques, including limiting your lines to 80 characters or
less. If you're using RStudio, go into Preferences \> Code \> Display to set up
a margin line to help you keep track of this guideline!

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(gridExtra)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
setwd("C:/Users/flfbarros/Documents/udacity/R/Red Wine Exploration")
redwine <- read.csv("wineQualityReds.csv", sep=",")
```

> **Tip**: Before you create any plots, it is a good idea to provide a short
introduction into the dataset that you are planning to explore. Replace this
quoted text with that general information!

This dataset is one of the datasets created on the paper "Modeling wine 
preferences by data mining from physicochemical properties" [Cortez et al., 
2009]. The inputs include objective tests, such as PH values, and the output is 
based on sensory data, gathered from at least 3 wine expert evaluations. The 
output value ranges from 0 (very bad) to 10 (very excellent). The wines are 
variants of the Portuguese "Vinho Verde" wine. This particular dataset contains 
information regarding the red wines.

There are 1599 different wines, 11 features and 1 output value (quality).
Information about the features and the output value is the following (provided 
by the original authors):

Input variables (based on physicochemical tests):

* 1 - fixed acidity (tartaric acid - g / dm^3)
* 2 - volatile acidity (acetic acid - g / dm^3)
* 3 - citric acid (g / dm^3)
* 4 - residual sugar (g / dm^3)
* 5 - chlorides (sodium chloride - g / dm^3
* 6 - free sulfur dioxide (mg / dm^3)
* 7 - total sulfur dioxide (mg / dm^3)
* 8 - density (g / cm^3)
* 9 - pH
* 10 - sulphates (potassium sulphate - g / dm3)
* 11 - alcohol (% by volume)

Output variable (based on sensory data):

* 12 - quality (score between 0 and 10)

Description of the features and output value is the following (also provided 
by the original authors):

* 1 - fixed acidity: most acids involved with wine or fixed or nonvolatile 
(do not evaporate readily)
* 2 - volatile acidity: the amount of acetic acid in wine, which at too high 
of levels can lead to an unpleasant, vinegar taste
* 3 - citric acid: found in small quantities, citric acid can add 'freshness' 
and flavor to wines
* 4 - residual sugar: the amount of sugar remaining after fermentation stops, 
it's rare to find wines with less than 1 gram/liter and wines with greater 
than 45 grams/liter are considered sweet
* 5 - chlorides: the amount of salt in the wine
* 6 - free sulfur dioxide: the free form of SO2 exists in equilibrium between 
molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial 
growth and the oxidation of wine
* 7 - total sulfur dioxide: amount of free and bound forms of S02; in low 
concentrations, SO2 is mostly undetectable in wine, but at free SO2 
concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine
* 8 - density: the density of water is close to that of water depending on the 
percent alcohol and sugar content
* 9 - pH: describes how acidic or basic a wine is on a scale from 0 (very 
acidic) to 14 (very basic); most wines are between 3-4 on the pH scale
* 10 - sulphates: a wine additive which can contribute to sulfur dioxide gas 
(S02) levels, wich acts as an antimicrobial and antioxidant
* 11 - alcohol: the percent alcohol content of the wine
* 12 - quality (score between 0 and 10)

# Univariate Plots Section

> **Tip**: In this section, you should perform some preliminary exploration of
your dataset. Run some summaries of the data and create univariate plots to
understand the structure of the individual variables in your dataset. Don't
forget to add a comment after each plot or closely-related group of plots!
There should be multiple code chunks and text sections; the first one below is
just to help you get started.

```{r echo=FALSE, message=FALSE}

str(redwine)

```


```{r echo=FALSE, Univariate_Plots}

summary(subset(redwine, select=-c(X)))

```

The dataset consists of 11 numerical features and 1 integer feature (quality). 
The X feature just represents the index of the observation on the data frame.

On the summary it is possible to identify some initial trends on the data. For 
example, there are some features that seem to have outliers on them, such as: 
volatile acidity, citric acid, redisual sugar, chlorides, free sulfur dioxide, 
total sulfur dioxide, and sulphates. Also, most of the wines are of average 
quality (the 1st quartile border equals 5, the median equals 6 and the third 
quartile border equals 6).

```{r echo=FALSE, message=FALSE, warning=FALSE}

table(redwine$quality)

```

Out of 1599 wines, 63 (~4%) are of quality 3 or 4, 1319 (~82.5%) are of quality 
5 or 6 and 217 (~13,6%) are of quality 7 or 8.

Histograms will help visualize distribution patterns for the other features
as well.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=quality)) +
  geom_histogram()

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=fixed.acidity)) +
  geom_histogram()

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=volatile.acidity)) +
  geom_histogram()

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=citric.acid)) +
  geom_histogram()

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=residual.sugar)) +
  geom_histogram()

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=chlorides)) +
  geom_histogram()

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=free.sulfur.dioxide)) +
  geom_histogram()

```

```{r  echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=total.sulfur.dioxide)) +
  geom_histogram()

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=density)) +
  geom_histogram()

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=pH)) +
  geom_histogram()

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=sulphates)) +
  geom_histogram()

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=alcohol)) +
  geom_histogram()

```

These histograms help us visualize how some of the features on the dataset have 
outliers, given that some of them are right tailed. Chlorides is a good example 
of this characteristic. Most of the data is around 0.05 and 0.15, with very 
few examples of wine going up to 6.

Also, most of the data is relatively normal, with the exception of citric acid.

Let's transform some of the features using a log scale to try and get better 
visualizations of them. Suitable choices would be residual sugar, chlorides, 
free sulfur dioxide, total sulfur dioxide, and sulphates.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=residual.sugar)) +
  geom_histogram() +
  scale_x_log10()

```


```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=chlorides)) +
  geom_histogram() +
  scale_x_log10()

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=free.sulfur.dioxide)) +
  geom_histogram() +
  scale_x_log10()

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=total.sulfur.dioxide)) +
  geom_histogram() +
  scale_x_log10()

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=sulphates)) +
  geom_histogram() +
  scale_x_log10()

```


With log transformation it is easier to identify patterns on these features. 
With the exception of free sulfur dioxide, they all have normal-like 
distributions.

> **Tip**: Make sure that you leave a blank line between the start / end of
each code block and the end / start of your Markdown text so that it is
formatted nicely in the knitted text. Note as well that text on consecutive
lines is treated as a single space. Make sure you have a blank line between
your paragraphs so that they too are formatted for easy readability.

# Univariate Analysis

> **Tip**: Now that you've completed your univariate explorations, it's time to
reflect on and summarize what you've found. Use the questions below to help you
gather your observations and add your own if you have other thoughts!

### What is the structure of your dataset?

The dataset consists of 1599 observations of red wines with information about 
11 physicochemical properties and 1 information about the quality of each wine. 
The pyhisicochemial information was collected with experiments and the quality 
of each wine is a mean of a value provided by different wine experts. All the 
information is in the form of numbers.

The quality values were given in a scale from 0 (very bad) to 10 (very 
excellent). However, the majority of the wines have quality of 5 or 6. The 
worst wine has a quality of 3. The best wine has a quality of 8.

Some features are long-tailed. Examples are residual sugar, chlorides, free 
sulfur dioxide, total sulfur dioxide, and sulphates.

With a few exceptions, the features have a distribution close to normal.

### What is/are the main feature(s) of interest in your dataset?

For the variance observed in some of these features and their nature in 
relation to overall taste in wine, I would describe residual sugar, density, 
and pH as features of interest. Of course, quality is the main focus of the 
dataset, so it is also of interest.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

The acidity properties (fixed acidity, volatile acidity, and citric acid) could 
support the information regarding pH. Alcohol is also a feature that could help 
predict the quality of a wine.

### Did you create any new variables from existing variables in the dataset?

I did not create any new variables from existing variables in the dataset.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

Some features (residual sugar, chlorides, free sulfur dioxide, total sulfur 
dioxide, and sulphates) are long-tailed, so it is difficult to analyze their 
distributions without any transformation. When adjusting them to a log 10 
scale, they become easier to visualize.

# Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

I've established some features of interest, but that was mainly done looking at 
their distribution and common wine knowledge. It would be best to evaluate the 
correlation between the features, to identify if the choices made are the best 
ones or if there is some possible feature worth considering.

```{r echo=FALSE, Bivariate_Plots}

cor(subset(redwine, select=-c(X)))

```

Based on correlations, the best features to predict the quality of a wine are, 
in order, alcohol, volatile acidity and sulphates. My initial intuition was 
that residual sugar, density and pH would better describe the quality of a 
wine, with alcohol and acidity features as possible features to look afterwards. 
These initial findings show that there is more to the quality of a wine than 
what common knowledge says.

With the correlation knowledge, I'll try now to visualize the influence of 
those features on the quality of the wines.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=alcohol, y=quality)) +
  geom_point(stat="summary", fun.y=mean) +
  geom_smooth()

```

The scatterplot suggests that as the percentage of alcohol increases so does 
the overall quality of the wine, although around 12-13% seems to be an ideal 
value.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(y=alcohol, x=as.factor(quality))) +
  geom_boxplot() +
  stat_summary(fun.y="mean", geom="point", shape=3, color="red") +
  xlab("quality")

```

The boxplot also supports that theory. Wines with better quality have higher 
median and higher mean (shown as red crosses).

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=volatile.acidity, y=quality)) +
  geom_point(stat="summary", fun.y=mean) +
  geom_smooth()

```

Volatile acidity is the opposite as alcohol: as it increases, the overall 
quality of the wine decreases. Since the correlation found was around -0.39, 
that was expected.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=as.factor(quality), y=volatile.acidity)) +
  geom_boxplot() +
  stat_summary(fun.y="mean", geom="point", shape=3, color="red") +
  xlab("quality")

```

Again, the boxplots suggests the same and provide additional information. As 
the level of volatile acidity decreases, the perceived quality of the wine 
increases.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=sulphates, y=quality)) +
  geom_point(stat="summary", fun.y=mean) +
  geom_smooth() + 
  scale_x_continuous(breaks=seq(0,2,0.1))

```

For sulphates, the scatterplot suggests that values around 0.75g/dm^3 and 
0.95g/dm^3 represents the best wines.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=as.factor(quality), y=sulphates)) +
  geom_boxplot() +
  stat_summary(fun.y="mean", geom="point", shape=3, color="red") +
  xlab("quality") +
  geom_hline(yintercept=0.75, linetype="longdash", alpha=0.5) +
  geom_hline(yintercept=0.95, linetype="longdash", alpha=0.5)

```

Following the best range for sulphates, established on the scatterplot, the 
boxplot shows that qualities of 7 or 8 have close to 50% of their wines with 
sulphates level on the specified range. Even though some qualities have some 
outliers on the range, most of their observations are below 0.75g/dm^3 of 
sulphates.

By plotting these three features on scatterplots and boxplots against quality, 
it is easier to visualize their influence on the quality of the wines.  

Since citric acid is the fourth most correlated feature to quality and it has 
some correlation with volatile acidity and sulphates, it is worth doing some 
analysis on it.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(y=citric.acid, x=volatile.acidity)) +
  geom_point(stat="summary", fun.y=mean) +
  geom_smooth() +
  scale_y_continuous(breaks=seq(-1,1,0.05))

```

The previous scatterplot between volatile acidity and quality suggested that 
good wines tend to have lower volatile adicity. This scatterplot between 
citric acid and volatile acidity shows that to obtain lower values of volatile 
acidity, the wine should have at least 0.35g/dm^3. I'll try that value against 
other features.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(y=citric.acid, x=sulphates)) +
  geom_point(stat="summary", fun.y=mean) +
  geom_smooth() +
  scale_y_continuous(breaks=seq(0,1,0.05)) +
  scale_x_continuous(breaks=seq(0.5,2,0.1)) +
  geom_vline(xintercept=0.75, linetype="longdash", alpha=0.5) +
  geom_vline(xintercept=0.95, linetype="longdash", alpha=0.5)

```

Since between 0.75 and 0.95 was the value range for sulphates on the best 
wines, I've added lines on the scatterplot to help identify the values of 
citric acid that fall on that range of sulphates. Again, 0.35g/dm^3 citric acid 
falls on the range of interest.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(y=citric.acid, x=alcohol)) +
  geom_point(stat="summary", fun.y=mean) +
  geom_smooth() +
  scale_y_continuous(breaks=seq(0,1,0.05)) +
  scale_x_continuous(breaks=seq(0,15,1)) +
  geom_vline(xintercept=12, linetype="longdash", alpha=0.5) +
  geom_vline(xintercept=13, linetype="longdash", alpha=0.5)

```

For alcohol, the range of interest is between 12% and 13%. Citric acid for 
those values are, on average, between 0.30g/dm^3 and 0.35g/dm^3.

When evaluating the correlation between citric acid and these other three 
features, it seems that 0.35g/dm^3 is the best value. Let's see if that holds 
true when comparing that feature with quality.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=citric.acid, y=quality)) +
  geom_point(stat="summary", fun.y=mean) +
  geom_smooth() +
  geom_vline(xintercept=0.35, linetype="longdash", alpha=0.5) +
  scale_x_continuous(breaks=seq(0,1,0.05))

```

As suggested by the relationship between other variables, 0.35g/dm^3 of citric 
acid indeed holds a good value of quality of the wine. However, the value is 
not the only one associated with a good wine. Other values also are correlated with good wine. 
Since the correlation between citric acid and quality was not very high 
(~0.226), that was expected.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=as.factor(quality), y=citric.acid)) +
  geom_boxplot() +
  stat_summary(fun.y="mean", geom="point", shape=3, color="red") +
  geom_hline(yintercept=0.35, linetype="longdash", alpha=0.5) +
  xlab("quality")

```

For the boxplot between citric acid and quality, we can see a clear different 
between better wines (quality of 7 and 8) and the other ones. They tend to have 
higher values of citric acid, as suggested in the previous scatterplot. The 
value of 0.35 alone doesn't add much information on the boxplot though.

Since density is a feature most commonly used when describing "the body" of the 
wine, I'll do a similar analysis for it.

```{r echo=FALSE, message=FALSE, warning=FALSE}

```


# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?



### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!