RED WINE QUALITY EXPLORATION by FELIPE BARROS
========================================================

> **Tip**: You will see quoted sections like this throughout the template to
help you construct your report. Make sure that you remove these notes before
you finish and submit your project!

> **Tip**: One of the requirements of this project is that your code follows
good formatting techniques, including limiting your lines to 80 characters or
less. If you're using RStudio, go into Preferences \> Code \> Display to set up
a margin line to help you keep track of this guideline!

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(gridExtra)
library(dplyr)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
setwd("C:/Users/felipe/Documents/Udacity/ND Data Science/R/RedWineEDA")
redwine <- read.csv("wineQualityReds.csv", sep=",")
```

> **Tip**: Before you create any plots, it is a good idea to provide a short
introduction into the dataset that you are planning to explore. Replace this
quoted text with that general information!

This dataset is one of the datasets created on the paper "Modeling wine 
preferences by data mining from physicochemical properties" [Cortez et al., 
2009]. The inputs include objective tests, such as PH values, and the output is 
based on sensory data, gathered from at least 3 wine expert evaluations. The 
output value ranges from 0 (very bad) to 10 (very excellent). The wines are 
variants of the Portuguese "Vinho Verde" wine. This particular dataset contains 
information regarding the red wines.

There are 1599 different wines, 11 features and 1 output value (quality).
Information about the features and the output value is the following (provided 
by the original authors):

Input variables (based on physicochemical tests):

* 1 - fixed acidity (tartaric acid - g / dm^3)
* 2 - volatile acidity (acetic acid - g / dm^3)
* 3 - citric acid (g / dm^3)
* 4 - residual sugar (g / dm^3)
* 5 - chlorides (sodium chloride - g / dm^3
* 6 - free sulfur dioxide (mg / dm^3)
* 7 - total sulfur dioxide (mg / dm^3)
* 8 - density (g / cm^3)
* 9 - pH
* 10 - sulphates (potassium sulphate - g / dm3)
* 11 - alcohol (% by volume)

Output variable (based on sensory data):

* 12 - quality (score between 0 and 10)

Description of the features and output value is the following (also provided 
by the original authors):

* 1 - fixed acidity: most acids involved with wine or fixed or nonvolatile 
(do not evaporate readily)
* 2 - volatile acidity: the amount of acetic acid in wine, which at too high 
of levels can lead to an unpleasant, vinegar taste
* 3 - citric acid: found in small quantities, citric acid can add 'freshness' 
and flavor to wines
* 4 - residual sugar: the amount of sugar remaining after fermentation stops, 
it's rare to find wines with less than 1 gram/liter and wines with greater 
than 45 grams/liter are considered sweet
* 5 - chlorides: the amount of salt in the wine
* 6 - free sulfur dioxide: the free form of SO2 exists in equilibrium between 
molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial 
growth and the oxidation of wine
* 7 - total sulfur dioxide: amount of free and bound forms of S02; in low 
concentrations, SO2 is mostly undetectable in wine, but at free SO2 
concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine
* 8 - density: the density of water is close to that of water depending on the 
percent alcohol and sugar content
* 9 - pH: describes how acidic or basic a wine is on a scale from 0 (very 
acidic) to 14 (very basic); most wines are between 3-4 on the pH scale
* 10 - sulphates: a wine additive which can contribute to sulfur dioxide gas 
(S02) levels, wich acts as an antimicrobial and antioxidant
* 11 - alcohol: the percent alcohol content of the wine
* 12 - quality (score between 0 and 10)

# Univariate Plots Section

> **Tip**: In this section, you should perform some preliminary exploration of
your dataset. Run some summaries of the data and create univariate plots to
understand the structure of the individual variables in your dataset. Don't
forget to add a comment after each plot or closely-related group of plots!
There should be multiple code chunks and text sections; the first one below is
just to help you get started.

```{r echo=FALSE, message=FALSE}

str(redwine)

```


```{r echo=FALSE, Univariate_Plots}

summary(subset(redwine, select=-c(X)))

```

The dataset consists of 11 numerical features and 1 integer feature (quality). 
The X feature just represents the index of the observation on the data frame.

On the summary it is possible to identify some initial trends on the data. For 
example, there are some features that seem to have outliers on them, such as: 
volatile acidity, citric acid, redisual sugar, chlorides, free sulfur dioxide, 
total sulfur dioxide, and sulphates. Also, most of the wines are of average 
quality (the 1st quartile border equals 5, the median equals 6 and the third 
quartile border equals 6).

```{r echo=FALSE, message=FALSE, warning=FALSE}

table(redwine$quality)

```

Out of 1599 wines, 63 (~4%) are of quality 3 or 4, 1319 (~82.5%) are of quality 
5 or 6 and 217 (~13,6%) are of quality 7 or 8.

Histograms will help visualize distribution patterns for the other features
as well.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=quality)) +
  geom_histogram()

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=fixed.acidity)) +
  geom_histogram()

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=volatile.acidity)) +
  geom_histogram()

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=citric.acid)) +
  geom_histogram()

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=residual.sugar)) +
  geom_histogram()

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=chlorides)) +
  geom_histogram()

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=free.sulfur.dioxide)) +
  geom_histogram()

```

```{r  echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=total.sulfur.dioxide)) +
  geom_histogram()

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=density)) +
  geom_histogram()

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=pH)) +
  geom_histogram()

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=sulphates)) +
  geom_histogram()

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=alcohol)) +
  geom_histogram()

```

These histograms help us visualize how some of the features on the dataset have 
outliers, given that some of them are right tailed. Chlorides is a good example 
of this characteristic. Most of the data is around 0.05 and 0.15, with very 
few examples of wine going up to 6.

Also, most of the data is relatively normal, with the exception of citric acid.

Let's transform some of the features using a log scale to try and get better 
visualizations of them. Suitable choices would be residual sugar, chlorides, 
free sulfur dioxide, total sulfur dioxide, and sulphates.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=residual.sugar)) +
  geom_histogram() +
  scale_x_log10()

```


```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=chlorides)) +
  geom_histogram() +
  scale_x_log10()

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=free.sulfur.dioxide)) +
  geom_histogram() +
  scale_x_log10()

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=total.sulfur.dioxide)) +
  geom_histogram() +
  scale_x_log10()

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=sulphates)) +
  geom_histogram() +
  scale_x_log10()

```


With log transformation it is easier to identify patterns on these features. 
With the exception of free sulfur dioxide, they all have normal-like 
distributions.

> **Tip**: Make sure that you leave a blank line between the start / end of
each code block and the end / start of your Markdown text so that it is
formatted nicely in the knitted text. Note as well that text on consecutive
lines is treated as a single space. Make sure you have a blank line between
your paragraphs so that they too are formatted for easy readability.

# Univariate Analysis

> **Tip**: Now that you've completed your univariate explorations, it's time to
reflect on and summarize what you've found. Use the questions below to help you
gather your observations and add your own if you have other thoughts!

### What is the structure of your dataset?

The dataset consists of 1599 observations of red wines with information about 
11 physicochemical properties and 1 information about the quality of each wine. 
The pyhisicochemial information was collected with experiments and the quality 
of each wine is a mean of a value provided by different wine experts. All the 
information is in the form of numbers.

The quality values were given in a scale from 0 (very bad) to 10 (very 
excellent). However, the majority of the wines have quality of 5 or 6. The 
worst wine has a quality of 3. The best wine has a quality of 8.

Some features are long-tailed. Examples are residual sugar, chlorides, free 
sulfur dioxide, total sulfur dioxide, and sulphates.

With a few exceptions, the features have a distribution close to normal.

### What is/are the main feature(s) of interest in your dataset?

Since quality is the output feature of the study that generated this dataset, 
the feature is undoubtly the main feature of interest.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

For the variance observed in some of these features and their nature in 
relation to overall taste in wine, I would describe residual sugar, density and 
pH as features of interest.

### Did you create any new variables from existing variables in the dataset?

I did not create any new variables from existing variables in the dataset.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

Some features (residual sugar, chlorides, free sulfur dioxide, total sulfur 
dioxide, and sulphates) are long-tailed, so it is difficult to analyze their 
distributions without any transformation. When adjusting them to a log 10 
scale, they become easier to visualize.

# Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

I've established some features of interest, but that was mainly done looking at 
their distribution and common wine knowledge. It would be best to evaluate the 
correlation between the features, to identify if the choices made are the best 
ones or if there is some possible feature worth considering.

```{r echo=FALSE, Bivariate_Plots}

cor(subset(redwine, select=-c(X)))

```

Based on correlations, the best features to predict the quality of a wine are, 
in order, alcohol, volatile acidity and sulphates. My initial intuition was 
that residual sugar, density and pH would better describe the quality of a 
wine, with alcohol and acidity features as possible features to look afterwards. 
These initial findings show that there is more to the quality of a wine than 
what common knowledge says.

With the correlation knowledge, I'll try now to visualize the influence of 
those features on the quality of the wines.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(y=alcohol, x=quality)) +
  geom_point(position=position_jitter(width=0.2, height=0.2)) +
  geom_smooth(method='lm')

```

The scatterplot suggests that as the percentage of alcohol increases so does 
the overall quality of the wine, as shown by the linear regression line. Even 
when comparing quality of 5 and 6 (the most common on the dataset), the 
observations tend to go up on quality 6, whilst being clumped below the 
regression line on quality 5.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(y=alcohol, x=as.factor(quality))) +
  geom_boxplot() +
  stat_summary(fun.y="mean", geom="point", shape=3, color="red") +
  xlab("quality")

```

The boxplot also supports that theory. Wines with better quality have higher 
median and higher mean (shown as red crosses).

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=volatile.acidity, y=quality)) +
  geom_point(position=position_jitter(width=0.2, height=0.2)) +
  geom_smooth(method='lm')

```

Volatile acidity is the opposite as alcohol: as it increases, the overall 
quality of the wine decreases. Since the correlation found was negative 
(around -0.39), that was expected.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=as.factor(quality), y=volatile.acidity)) +
  geom_boxplot() +
  stat_summary(fun.y="mean", geom="point", shape=3, color="red") +
  xlab("quality")

```

Again, the boxplots suggests the same and provide additional information (mean
and median values for each quality). As the level of volatile acidity decreases,
the perceived quality of the wine increases.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(y=sulphates, x=quality)) +
  geom_point(position=position_jitter(width=0.2, height=0.1)) +
  geom_smooth(method='lm')

```

For sulphates, the scatterplot has a behavior similar to alcohol: better 
quality wines have higher levels of sulphates.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=as.factor(quality), y=sulphates)) +
  geom_boxplot() +
  stat_summary(fun.y="mean", geom="point", shape=3, color="red") +
  xlab("quality")

```

The boxplot for each quality supports that trend. All the higher qualities have 
a higher mean and median values of sulphates.

By plotting these three features on scatterplots and boxplots against quality, 
it is easier to visualize their influence on the quality of the wines. By now, 
it is safe to say that better wines probably have higher alcohol and sulphates 
levels, whilst having a lower volatile acidity level.

Since citric acid is the fourth most correlated feature to quality, it is worth 
doing some analysis on it.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(y=citric.acid, x=quality)) +
  geom_point(position = position_jitter(width=0.2, height=0.1)) +
  geom_smooth(method='lm')

```

The correlation between citric acid and quality (~0.226) is not much different 
from the correlation between sulphates and quality (~0.25), so their tendencies 
are also similar to each other. Better wines tend to have, on average, higher 
levels of citric acid.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=as.factor(quality), y=citric.acid)) +
  geom_boxplot() +
  stat_summary(fun.y="mean", geom="point", shape=3, color="red") +
  xlab("quality")

```

For the boxplot between citric acid and quality, we can see a clear difference
between better wines (quality of 7 and 8) and the other ones. Not only they 
tend to have higher values of citric acid, the majority of their observations 
(over 75%) are over 0.25g/dm^3 of citric acid.

Since density is a feature most commonly used when describing "the body" of the 
wine, I'll do a similar analysis for it.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=quality, y=density)) +
  geom_point(position=position_jitter(width=0.2, height=0.01)) +
  geom_smooth(method='lm')

```

Since the correlation between these features is very low (around -0.17), the 
influence of density on the quality of the wine is barely noticeable. It seems 
that using density to evaluate the quality of a wine (a common practice of 
non-experts) is not the best approach.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=as.factor(quality), y=density)) +
  geom_boxplot() +
  stat_summary(fun.y="mean", geom="point", shape=3, color="red") +
  xlab("quality")

```

The boxplot adds a little more information. There is a decrease of density 
levels on better wines, but that is only noticeable with quality levels 
of 7 and 8. For the other quality levels, there doesn't seem to be a 
significant difference.

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

I've investigated the relationships between quality and alcohol, volatile 
acidity, sulphates, citric acid and density. In terms of correlation with 
quality, they were, in order, 1st, 2nd, 3rd, 4th and 6th most correlated 
features of the dataset. The first four were chosen because of they closer 
relationship with quality and density was chosen out of personal curiosity.

Alcohol was the feature most correlated with quality, with a factor of 0.48. 
The positive correlation between them implied that increasing the level of 
alcohol would increase the perceived quality of the wine. The plots created 
demonstrated that assumption.

Volatile acidity was the second feature most correlated with quality, with a 
factor of -0.39. The negative correlation implies a decrease of one when an 
increase in the other occurs. The plots generated help visualize that 
relationship. As volatile acidity levels go down, the quality of the wines go 
up.

Sulphates and citric acid where the third and fourth most correlated qualities, 
with factors of 0.25 and 0.226, respectively. Higher levels of these features 
are correlated with better quality of the wine, although their influence is not 
as strong as the influence of alcohol and volatile acidity.

Density was the sixth most correlated feature with quality. The choice to 
visualize the influence of that feature was mostly curiosity based, since the 
correlation factor is of only -0.175. Although there is an influence of the 
feature on the quality of the wine, that influence is only noteworthy on the 
best wines on the dataset (7 or above). Those wines tend to have a lower 
density value.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

Features related to acidity naturally seem to be correlated with each other. 
Examples are fixed acidity, volatile acidity, citric acid and pH. Density also 
seems to be related with those features, along with residual sugar and alcohol.

### What was the strongest relationship you found?

The strongest relationship found was between fixed acidity and pH, with a 
factor of 0.683. That is not a surprising relationship, as one feature (pH) is 
evaluated based on properties of the other (acidity).

# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

To help visualization for the next plots, I'll divide the wines on baskets, 
based on their quality. The baskets are:

* Wines with quality below 5
* Wines with quality exactly 5
* Wines with quality exactly 6
* Wines with quality above 6

The choice of breaks for the baskets are based on the fact that most 
observations on the dataset have quality of 5 or 6.

```{r echo=FALSE, Multivariate_Plots}

redwine$quality_factor <- cut(redwine$quality, breaks=c(0,4,5,6,10))

```

```{r echo=FALSE, Multivariate_Plots}

table(redwine$quality_factor)

```

By reducing quality factor levels from 6 (3-8) to 4, visualization on the next 
plots is made clearer.


```{r echo=FALSE, Multivariate_Plots}

ggplot(data=redwine, aes(x=alcohol, y=volatile.acidity, 
                         color=quality_factor)) +
  geom_point(position=position_jitter(width=0.2, height=0.1), alpha=1/3) +
  geom_smooth(method='lm', alpha=0) +
  scale_y_continuous(breaks=seq(0,1.5,0.1))

```

This plot reveal a lot of information. It was already known by previous 
plots that better wines have lower volatile acidity levels. Not only this plot 
corroborate that, it also correlates that with the information that better 
wines tend to have higher alcohol levels. The worst wines (red line) not only 
have the highest mean value for volatile acidity (being above the rest of the 
lines), they also tend to increase volatile acidity as they increase alcohol 
levels. It seems that, as they try to solve one issue (by increasing alcohol 
levels), they create another (by increasing volatile acidity levels). Wines 
from other quality baskets tend to maintain or lower their acidity levels when 
increasing the alcohol levels, with the exception of the higher quality basket, 
that increased the volatile acidity by a minor margin when increasing their 
alcohol levels.


```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=alcohol, y=density, color=quality_factor)) +
  geom_point(position=position_jitter(width=0.2, height=0.005)) +
  geom_smooth(method='lm', alpha=0)

```

When trying to do the same with alcohol, density and quality factor, it is 
possible to see a even more distinct pattern. Wines with less alcohol tend to 
be more dense. Previous plots showed that higher levels of alcohol and lower 
levels of density were correlated with better quality of the wine. This plot 
suggests that all the baskets of quality follow the trend of trying to increase 
the alcohol levels whilst also trying to decrease the density levels.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(data=redwine, aes(x=volatile.acidity, y=sulphates, color=quality_factor)) +
  geom_point(position=position_jitter(width=0.1, height=0.1)) +
  geom_smooth(method='lm', alpha=0)

```

This is a very good plot to demonstrate how these features influence the 
quality of a wine. There are two very distinct masses demonstrated here. The 
bad and average wines (red and green, respectively) are oriented towards the 
southeast section of the graph. The above average and good wines (blue and 
purple, respectively) are oriented towards the northwest section of the graph. 
Of course a few outliers are present, but the majority of the wines of each 
basket follow that behavior.

So far, the only categorical feature in the dataset was quality, that was 
reduced into buckets. I've noticed that having almost all the variables with 
continuous values might conceal some interesting information. For example, is 
there a difference in how the density affects the quality of the wine in 
different levels of alcohol?

I'll bucket wines by alcohol levels and investigate some more.

```{r echo=FALSE, message=FALSE, warning=FALSE}

summary(redwine$alcohol)

```

A reminder of the summary for alcohol values on the dataset. The value range is 
of 6.5 (14.9 minues 8.4). By dividing alcohol levels into three buckets, that 
would imply in a step of around 2.16 for each bucket. I'll round it to 2.15 and 
use that value as a step to the cut. The buckets will be cut as follows:

* Low alcohol level: over 8.4% or under or equal to 9.55%
* Medium alcohol level: over 9.55% or under or equal to 11.7%
* High alcohol level: over 11.7%

```{r echo=FALSE, message=FALSE, warning=FALSE}

redwine$alcohol_level <- cut(redwine$alcohol, breaks=c(8,9.55,11.7,15))
table(redwine$alcohol_level)
```

The dataset is mostly of medium alcohol levels, based on the cut parameters. 
However, there is a reasonable distribution for the other buckets.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!